Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-15 doc/mc-paper.tex
Lots of updates

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\drational}{\mathbb{Q}_{2,5}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\generator}{\progname{d2p}}
\newcommand{\Generator}{\progname{D2p}}
\newcommand{\checker}{\progname{crat-check}}
\newcommand{\Checker}{\progname{Crat-check}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}

@@

We have developed a program that can generate POG representations
from the d-DNNF graphs generated by the state-of-the-art knowledge
compiler \dfour, as well as independently checkable proofs that the
output POGs are equivalent to the input CNF formulas.
We have implemented an efficient checker
that can handle very large proofs. Our tool
chain for generating and verifying POGs scales to all but the largest d-DNNF graphs
generated by \dfour for formulas
from a recent model counting competition. Moreover, we
present a checker that has been formally verified
in the \lean{} proof assistant, thereby creating the
first formally verified toolchain for
% weighted and unweighted
model counting.
\end{abstract}

@@

Modern Boolean satisfiability (SAT) solvers provide the ability to
find a satisfying assignment to a Boolean formula or to prove that no
such assignment exists.  The have applications across a variety of
domains, including formal verification of hardware, software, and
security protocols, computational mathematics, and combinatorial
optimization.  Some problems of interest, however, require going
beyond standard Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments, often in cases where there are far to many satisfying
assignments to enumerate individually.  Model counting, in turn, has
applications in artificial intelligence, computer security, and
statistical sampling.  A variant on standard model counting is {\em
  weighted model counting}, where a weighting function is defined for
each assignment and the goal becomes to compute the sum of the weights
of the satisfying assignments.

@@

One approach to model counting, known as {\em knowledge compilation}
is to transform the formula into a structured form for which model
counting becomes more tractable.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a Boolean formula as a
directed acyclic graph, with terminal nodes labeled by Boolean
variables and their complements, and with each nonterminal node
labeled by Boolean operation And or Or.  Restrictions are placed on the
structure of the graph such that a count of the models can be computed
by a single bottom-up traversal of the graph.

Once a Boolean formula has been transformed into a d-DNNF
representation, a number of useful properties of the formula can
readily be computed.  Kimmig, et al., present a very general
framework, known as {\em algebraic model
counting}~\cite{kimmig:jal:2017}, that go well beyond simple model counting.

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are corrected, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers are able to produce
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has
shown that adding this capability has also greatly improved the quality of
SAT solvers and the ability of developers to quickly detect and
identify bugs in their solvers.

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations commonly used in knowledge
compilation, including d-DNNF\@.  We present the CRAT file format that
both captures the structure of a POG and can provide a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof is expressed as a sequence of
clause addition and deletion steps, based on the extended resolution
proof system~\cite{Tseitin:1983}.

We have developed a program that can generate CRAT descriptions from
the d-DNNF graphs generated by the state-of-the-art knowledge compiler
\dfour, as well as a fast CRAT proof checker.  We evaluate the
capabilities of our tool chain using benchmark formulas from the 2022
standard and weighted model competitions.  Our tools are able to
handle all but the largest d-DNNF graphs generated by \dfour.
Moreover, we
present a CRAT checker, and a POG model counter that have been formally verified
in the \lean{} proof assistant, thereby creating the
first formally verified toolchain for
% weighted and unweighted
model counting.

@@

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{cruz-cade-2017,RAT,wetzler14_drattrim}.  Some progress has also
been made on extending these proof systems to quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  We compare the performance of
our tool chain to theirs as part of our experimental evaluation in
Section~\ref{sect:experimental}.  Our experience with developing a
formally verified proof system has shown us that, even for the well
understood framework of extended resolution, ensuring that the system
is sound can be challenging.

@@

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
  normal form, and 2) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses in a formula and the
  literals in a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

@@

  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, the summation $\sum$ is computed using addition operation $\radd$, and the product $\prod$ is computed using multiplication operation $\rmul$.
\label{def:labeling}
\end{definition}

Many important queries about Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi, w)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a randomized test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

Generalizations of this framework are discussed in Section~\ref{sect:future}.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.  A nonterminal node is labeled by one of three Boolean operations:
$\boolnot$, $\pand$, or $\por$.  Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.

As can be seen, a POG is simply a way to represent a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
$\phi(\nodeu)$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi(\nodeu))$.

We define the {\em size} of POG $P$, written $|P|$ to be the sum of
the number of nonterminal nodes and the number of edges.  Ring
evaluation of $P$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
The final result is then $\rep(\noder, w)$.

@@

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

\section{The CRAT POG Representation and Proof System}

\begin{figure}
\centering{
\begin{tikzpicture}[scale=0.18]
  \draw (8,10) rectangle (20,18);
  \node at (14,15.2) {Knowledge};
  \node at (14,12.8) {Compiler};

  \draw (48,10) rectangle (60,18);
  \node at (54,15.2) {Proof};
  \node at (54,12.8) {Checker};

  \draw (48,0) rectangle (60,8);
  \node at (54,5.2) {Ring};
  \node at (54,2.8) {Evaluator};

  %% CNF
  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (48,16);
  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
  \node [left] at (-1,20) {\large {$\phi$}};
  \node [above] at (2,20) {\texttt{.cnf}};

  %% CRAT
  \draw[thick] (40,12) [-latex] -- (48,12);
  \draw[thick] (44,12) -- (44,4) [-latex] -- (48,4);
  \node [above] at (44,12) {\texttt{.crat}};

  %% OK/Not
  \draw[thick] (60,14) [-latex] -- (64,14) ;
  \node [right] at (64,15.2) { OK /} ;
  \node [right] at (64,12.8) { Not OK} ;
  \draw[thick] (60,4) [-latex] -- (64,4) ;
  \node [right] at (64,4) {\large {$\textbf{R}(\phi)$}};

\end{tikzpicture}
}
\caption{Tool Chain. The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

A CRAT file provides both a declaration of a POG, as well as an easily checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG, thus enabling certified
knowledge compilation.  The proof system is based on extended
resolution~\cite{Tseitin:1983} with reverse unit propagation
(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
proving that a set of clauses logically implies another clause.

As is shown in Figure~\ref{fig:chain}
two programs are involved in producing a certified compilation of a formula:
\begin{itemize}
\item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
\item The {\em proof checker} ensures that all of the proof conditions are satisfied.
\end{itemize}

@@

Figure~\ref{fig:chain} illustrates one method of generating a CRAT
file from an input formula in conjunctive normal form.  Our tool chain
is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
but it would also work for other tools that use a top-down search
strategy to generate a d-DNNF representation.  Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every d-DNNF node explicitly,
our program performs simplifications
to eliminate Boolean constants.  Except for the degenerate cases where the formula is unsatisfiable or a tautology, we can
therefore assume that the POG contains only product, sum,
and literal nodes.

@@

For input formula $\phi$ and its translation into a POG with root literal
$r$, the most challenging part of the proof is to show that
$\modelset(\phi) \subseteq \modelset(P)$, i.e., that any total
assignment $\alpha$ that is a model of $\phi$ will yield $\assign(r) =
1$, for root literal $r$.
This part of the proof consists of a series of clause assertions
leading to one adding $r$ as a unit clause.

>>>>

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
\newcommand{\drational}{\mathbb{Q}_{2,5}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}

@@

We have developed a program that can generate POG representations from
the d-DNNF graphs generated by the state-of-the-art knowledge compiler
\dfour, as well as independently checkable proofs that the output POGs
are equivalent to the input CNF formulas.  We have implemented an
efficient checker that can handle very large proofs. Our tool chain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs generated by \dfour{} for formulas from a recent model counting
competition. Moreover, we present a checker and a model counter for POGs that have
been formally verified in the \lean{} proof assistant, thereby
creating the first formally verified toolchain for
weighted and unweighted
model counting.
\end{abstract}

@@

Modern Boolean satisfiability (SAT) solvers provide the ability to
find a satisfying assignment to a Boolean formula or to prove that no
such assignment exists.  They have applications across a variety of
domains, including formal verification of hardware, software, and
security protocols, computational mathematics, and combinatorial
optimization.  Some problems of interest, however, require going
beyond standard Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting, in turn, has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

@@

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
becomes straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by Boolean operation And or Or.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal of the graph.
Kimmig, et al., present a very general framework, known as {\em
  algebraic model counting}~\cite{kimmig:jal:2017}, describing
properties of Boolean functions that can be efficiently computing from
its d-DNNF representation.  These include standard and weighted model
counting, and much more.

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that having this capability gives developers the ability quickly
detect and identify bugs in their solvers, and this, in turn, has led
to more reliable SAT solvers.

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations commonly used in knowledge
compilation, including d-DNNF\@.  We present the CRAT file format.
It captures both the structure of a POG and a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof consists of a sequence of
clause addition and deletion steps, based on the extended resolution
proof system~\cite{Tseitin:1983}.  We present a set of conditions that
a CRAT file must satisfy to provide an equivalence proof.

\begin{figure}
\centering{
\begin{tikzpicture}[scale=0.17]
  \draw (8,10) rectangle (20,18);
  \node at (14,15.2) {Knowledge};
  \node at (14,12.8) {Compiler};

  \draw (50,10) rectangle (62,18);
  \node at (56,15.2) {Proof};
  \node at (56,12.8) {Checker};

  \draw (50,0) rectangle (62,8);
  \node at (56,5.2) {Ring};
  \node at (56,2.8) {Evaluator};

  \draw[dashed] (48,-2) rectangle (64,22);
  \node at (56,20) {Trusted Code};

  %% CNF
  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
  \node [left] at (-1,20) {\large {$\phi$}};
  \node [above] at (2,20) {\texttt{.cnf}};

  %% CRAT
  \draw[thick] (40,12) [-latex] -- (50,12);
  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
  \node [above] at (44,12) {\texttt{.crat}};

  %% OK/Not
  \draw[thick] (62,14) [-latex] -- (66,14) ;
  \node [right] at (66,15.2) { OK /} ;
  \node [right] at (66,12.8) { Not OK} ;
  \draw[thick] (62,4) [-latex] -- (66,4) ;
  \node [right] at (66,4) {\large {$\textbf{R}(\phi)$}};

\end{tikzpicture}
}
\caption{Certifying Tool Chain.
  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
and (weighted) model counting tool chain that we have developed.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph generated by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this tool chain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}. 
As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
shifts the need for trustworthy code away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trustworthy---its errors will also be caught by the proof checker.

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant.  Running these two programs
has the effect of formally verifying both the logical equivalence of the POG to the input formula and the correctness of the count.
In addition, the
existence of a formally verified checker serves the important role of
establishing the soundness of our proof framework.

@@

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,wetzler14_drattrim,Tan:2021}.  Some progress has also
been made on extending these proof systems to quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.  There tool
chain can be applied to some programs that are pure model counters,
while ours can be applied to pure knowledge compilers.  Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Section~\ref{sect:experimental}.  Our
experience with developing a formally verified CRAT proof checker has shown
us that, even for the well understood framework of extended
resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.

@@

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em negation normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
  negation normal form, and 2) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses in a formula and the
  literals in a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

@@

  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum $\sum$ is computed using addition operation $\radd$, and product $\prod$ is computed using multiplication operation $\rmul$.
\label{def:labeling}
\end{definition}

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi, w)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
For any labeling $w$, ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies:
\begin{eqnarray}
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\rep\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\rep\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
As can be seen, a POG is simply a way to represent a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

We define the {\em size} of POG $P$, written $|P|$ to be the 
the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
evaluation of $P$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
The final result is then $\rep(\noder, w)$.

@@

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

\section{Clausal Proof Frameworks}

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step maintaining logical equivalence to
the original formula.  More precisely, the status of the proof at any step will be represented by
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables}.  Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$, 
we want to maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  We will use
this form only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  The other is by {\em
  implication redundancy}.  That is, we can add clause $C$ to the set
of clauses forming formula $\phi$ when $\phi \imply C$.  Clauses can
also be deleted through implication redundancy.  That is, given a set
of clauses $\{C\} \cup \phi$, we can remove $C$ if $\phi \imply C$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic.
It is a way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_p$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

Compared to the proofs of unsatisfiability generated by SAT solvers,
it can be seen that ours has important differences.  Most
significantly, we want our proofs to maintain logical equivalence at
each step, whereas an unsatisfiability proof need only guarantee that
no proof step makes a satisfiable set of clauses become
unsatisfiable.  Therefore, our proofs must justify clause deletions as well as clause additions.

\section{The CRAT POG Representation and Proof System}
\label{sect:crat}

A CRAT file provides both a declaration of a POG, as well as an easily checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG, thus enabling certified
knowledge compilation.
%% The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.

@@

Figure~\ref{fig:chain} illustrates one method of generating a CRAT
file from an input formula in conjunctive normal form.  Our tool chain
is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
but it would also work for other tools that use a top-down search
strategy to generate a d-DNNF representation.  Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except for the degenerate cases
where the formula is unsatisfiable or a tautology, we can also assume
that the POG does not contain any constant nodes.
In addition, negation operators will
only be applied to variables, and so we can view the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

For input formula $\inputformula$ and its translation into a POG with root literal
$r$, the most challenging part of the proof is to show that
$\modelset(\inputformula) \subseteq \modelset(P)$, i.e., that any total
assignment $\alpha$ that is a model of $\inputformula$ will yield $\assign(r) =
1$, for root literal $r$.
This part of the proof consists of a series of clause assertions
leading to one adding $r$ as a unit clause.
